---
title: "Practice Sheet Day 2"
subtitle: "2nd SARA Statistics Winter School"
author: "Dr. Ajay Koli"
date: today
format: html
toc: true
number-sections: true
---

## Frequency Distribution

### Step 1: Create a Data Set

```{r}
# Read a CSV file
dating <- read.csv("data/dating.csv")

# Display the first few rows
head(dating)
```

Always check the structure of the imported data:

```{r}
str(dating)
summary(dating)
```

### Step 2: Frequency Table

We first calculate the frequency of each score:

```{r}
freq_table <- table(dating$HighSalary)

print(freq_table)
```

### Step 3: Calculate Relative Frequencies

Divide each frequency by the total number of observations:

```{r}
# Total observations
total_observations <- sum(freq_table)

# Relative frequency
relative_freq <- (freq_table / total_observations)*100
print(relative_freq)
```

This means:

- 5% of ladies chose "score 1" for high salary
    
- 20% of ladies chose "score 4" for high salary

- 10% of ladies chose "score 10" for high salary

### Step 4: Visualizing Relative Frequencies using Bar Plot.

```{r}
# Bar plot
barplot(relative_freq, col = rainbow(10), main = "Relative Frequency of Favorite Fruits", 
        xlab = "Fruit", ylab = "Relative Frequency")

```

### Frequency Polygon

1. Generate or Load Data

We will create an example dataset of students' exam scores:

```{r}
# Example dataset: Exam scores of 50 students
set.seed(123)
exam_scores <- round(rnorm(50, mean = 75, sd = 10), 0)

exam_scores
```

2. Create a Frequency Table

Define intervals (bins) and calculate frequencies:

```{r}
# Define bins (class intervals)
bins <- seq(50, 100, by = 5)  # Intervals from 50 to 100 with a width of 5

# Create a frequency table
freq_table <- table(cut(exam_scores, breaks = bins, right = FALSE))
print(freq_table)

```

3. Plot the Frequency Polygon

Extract the midpoints of each class interval and use them for the x-axis:

```{r}
# Calculate midpoints of the bins
midpoints <- (bins[-1] + bins[-length(bins)]) / 2

# Frequency values
freq_values <- as.numeric(freq_table)

# Plot the frequency polygon
plot(midpoints, freq_values, type = "o", col = "blue", lwd = 2,
     xlab = "Exam Scores", ylab = "Frequency", 
     main = "Frequency Polygon of Exam Scores")

# Add grid for better visualization
grid()

```

4. Adding a Histogram for Comparison

You can overlay a histogram with the frequency polygon:

```{r}
# Plot histogram
hist(exam_scores, breaks = bins, freq = TRUE, col = "lightgray", 
     main = "Frequency Polygon Overlaid on Histogram", xlab = "Exam Scores")

# Overlay frequency polygon
lines(midpoints, freq_values, type = "o", col = "red", lwd = 2)

```

## Normal Distribution in R

### 1. Simulate Data

Generate random data following a normal distribution:

```{r}
# Generate 1000 random values with mean = 50 and standard deviation = 10
set.seed(123)
data <- rnorm(1000, mean = 50, sd = 10)
```

### 2. Plot the Normal Distribution

Visualize the data with a histogram and overlay the theoretical normal curve:

```{r}
# Histogram of the data
hist(data, breaks = 30, prob = TRUE, col = "lightblue", 
     main = "Normal Distribution with Mean and SD", xlab = "Values")

# Overlay the normal curve
curve(dnorm(x, mean = mean(data), sd = sd(data)), 
      col = "red", lwd = 2, add = TRUE)
```

### 3. Central Tendency and Variability

Compute the mean, median, and standard deviation:

```{r}
# Central tendency
mean_value <- mean(data)
median_value <- median(data)

# Variability
sd_value <- sd(data)

# Print the results
cat("Mean:", mean_value, "\nMedian:", median_value, "\nStandard Deviation:", sd_value, "\n")
```

### Visualizing Central Tendency

Add vertical lines for the mean, median, and ±1 standard deviation:

```{r}
# Histogram with indicators
hist(data, breaks = 30, prob = TRUE, col = "lightblue", 
     main = "Normal Distribution with Central Tendency", xlab = "Values")

# Overlay normal curve
curve(dnorm(x, mean = mean(data), sd = sd(data)), 
      col = "red", lwd = 2, add = TRUE)

# Add mean, median, and standard deviation lines
abline(v = mean_value, col = "gold", lwd = 2, lty = 2)  # Mean
abline(v = median_value, col = "green", lwd = 2, lty = 2)  # Median
abline(v = mean_value + sd_value, col = "purple", lwd = 2, lty = 3)  # +1 SD
abline(v = mean_value - sd_value, col = "purple", lwd = 2, lty = 3)  # -1 SD
legend("topright", legend = c("Mean", "Median", "+1 SD", "-1 SD"),
       col = c("gold", "green", "purple", "purple"), lty = c(2, 2, 3, 3), lwd = 2)

```

### 5. Interpret the Results

- Shape: The histogram matches the bell shape of a normal distribution.

- Mean and Median: Both align, confirming symmetry.

- Spread: The standard deviation defines how data points are distributed around the mean.

## Skewness & Kurtosis

### 1. Simulating Data

Create datasets with different skewness and kurtosis:

```{r}
# Libraries
library(e1071)  # For skewness
library(moments)  # For kurtosis

# Data with normal distribution
set.seed(123)
data_normal <- rnorm(1000, mean = 50, sd = 10)

# Data with positive skew
data_pos_skew <- rexp(1000, rate = 0.1)

# Data with negative skew
data_neg_skew <- -rexp(1000, rate = 0.1) + 100

```

### 2. Calculating Skewness and Kurtosis

```{r}
# Skewness
cat("Normal Data Skewness:", skewness(data_normal), "\n")
cat("Positive Skew Data Skewness:", skewness(data_pos_skew), "\n")
cat("Negative Skew Data Skewness:", skewness(data_neg_skew), "\n")

# Kurtosis
cat("Normal Data Kurtosis:", kurtosis(data_normal), "\n")
cat("Positive Skew Data Kurtosis:", kurtosis(data_pos_skew), "\n")
cat("Negative Skew Data Kurtosis:", kurtosis(data_neg_skew), "\n")

```

### 3. Visualizing Distributions

```{r}
# Plot all distributions
#par(mfrow = c(3, 1))  # Arrange plots in a single column

# Normal Distribution
hist(data_normal, breaks = 30, prob = TRUE, col = "lightblue", main = "Normal Distribution")
curve(dnorm(x, mean = mean(data_normal), sd = sd(data_normal)), col = "red", lwd = 2, add = TRUE)

# Positive Skew
hist(data_pos_skew, breaks = 30, prob = TRUE, col = "lightgreen", main = "Positive Skew Distribution")

# Negative Skew
hist(data_neg_skew, breaks = 30, prob = TRUE, col = "lightcoral", main = "Negative Skew Distribution")

```

## Example for Skewness and Kurtosis

### 1. Generate Custom Data

We will create datasets with varying skewness and kurtosis:

```{r}
# Libraries
library(e1071)   # For skewness
library(moments) # For kurtosis

# Custom datasets
set.seed(456)
data_normal <- rnorm(1000, mean = 50, sd = 10)       # Normal distribution
data_pos_skew <- rexp(1000, rate = 0.1)             # Positive skew
data_neg_skew <- -rexp(1000, rate = 0.1) + 100      # Negative skew
data_high_kurt <- c(rnorm(950, mean = 50, sd = 5),  # High kurtosis
                    rnorm(50, mean = 50, sd = 25))
data_low_kurt <- runif(1000, min = 30, max = 70)    # Low kurtosis

```

### 2. Calculate Skewness and Kurtosis

```{r}
# Skewness and Kurtosis summary
results <- data.frame(
  Dataset = c("Normal", "Positive Skew", "Negative Skew", "High Kurtosis", "Low Kurtosis"),
  Skewness = c(skewness(data_normal), skewness(data_pos_skew), skewness(data_neg_skew), 
               skewness(data_high_kurt), skewness(data_low_kurt)),
  Kurtosis = c(kurtosis(data_normal), kurtosis(data_pos_skew), kurtosis(data_neg_skew), 
               kurtosis(data_high_kurt), kurtosis(data_low_kurt))
)
print(results)

```

### 3. Annotated Visualizations

```{r}
# Plot all distributions
#par(mfrow = c(3, 2))  # Arrange plots in a 3x2 grid

# Function to annotate skewness and kurtosis on a histogram
plot_distribution <- function(data, title) {
  hist(data, breaks = 30, prob = TRUE, col = "lightblue", main = title, xlab = "Values")
  skew <- round(skewness(data), 2)
  kurt <- round(kurtosis(data), 2)
  text(x = mean(data), y = 0.02, 
       labels = paste("Skewness:", skew, "\nKurtosis:", kurt), col = "red")
}

# Plot distributions with annotations
plot_distribution(data_normal, "Normal Distribution")
plot_distribution(data_pos_skew, "Positive Skew Distribution")
plot_distribution(data_neg_skew, "Negative Skew Distribution")
plot_distribution(data_high_kurt, "High Kurtosis Distribution")
plot_distribution(data_low_kurt, "Low Kurtosis Distribution")

```

### Advanced Analysis

**Boxplots for Comparison:**

```{r}
# Combine all datasets into a data frame with proper grouping
all_data <- data.frame(
  Values = c(data_normal, data_pos_skew, data_neg_skew, data_high_kurt, data_low_kurt),
  Category = rep(c("Normal", "Positive Skew", "Negative Skew", "High Kurtosis", "Low Kurtosis"), 
                 each = 1000) # Correct grouping
)

# Create a boxplot for comparison
boxplot(Values ~ Category, data = all_data, col = rainbow(5),
        main = "Comparison of Distributions via Boxplot", 
        ylab = "Values", xlab = "Distribution Types")

```






